An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2021-12-27 simplified_sdc_intro/paper.tex

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence.  However, for very strong reactions, when using
Strang splitting the state can drift significantly off of the smooth
solution to the coupled reactive hydrodynamics equations, as shown
graphically in \cite{astronum:2018} using an earlier version of the
present algorithm.

A variation on Strang splitting called (re-)balanced splitting was
developed in \citet{speth:2013}.

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity)
as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep: 
\begin{align}
\label{eq:dt:nuce}
\Delta t &\le f_e\, \min_{i} \left\{\frac{e_{i}}{\dot{S}_{i}}\right\} \\
%
\label{eq:dt:nucX}
\Delta t &\le f_X\, \min_{i} \left\{\min_{k,X_k > \epsilon_X}\frac{{X_k}_{i}}{{\omegadot(X_k)_i}}\right\}
\end{align}
where $i$ is the zone index and $f_e$ and $f_X$ are runtime parameters
used to control the allowed change, and only species for which $X_k > \epsilon_X$ are considered.

>>>>

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence and showed that integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  A variation on Strang splitting called (re-)balanced
splitting was developed in \citet{speth:2013}, but we do not explore
that here. \MarginPar{keep this sentence?}

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity)
as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep.  An energy limiter takes the form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{\int \dot{S}_i(t) dt} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Eq.~\ref{eq:dt:nuce_I} uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Eq.~\ref{eq:dt:nuce_II} uses the integral of the
energy generation rate over the entire step (or last Strang half).
